---
title: "Analyze bulk IEC Piebald RNAseq data"
author: "Lev Litichevskiy, adjusted for this manuscript by Megan Liou"
date: "December 15, 2022/Updated Aug 26, 2026"
urlcolor: blue
---

# Load packages
```{r, warning=F, message=F}
library(tidyverse)
library(dplyr)
library(DESeq2)
library(ggrepel)
#library(clusterProfiler)
library(org.Mm.eg.db)

theme_set(theme_bw(base_size=18))

```

# Import metadata

```{r}
meta.df <- read.csv("~/Documents/piebald/piebald_RNAseq_metadata_221209.csv")
#meta_slsl.df = subset(meta.df, genotype == "slsl")

counts.df <- read.table("~/Documents/piebald/piebald_kallisto_counts_matrix.tsv", sep="\t", header=T, check.names = F) %>% 
column_to_rownames("gene.name")
```

# Convert to integer
```{r}
# round to the nearest integer
counts.mat.int <- counts.df %>% 
  as.matrix() %>% 
  round()

# convert data type to integer
mode(counts.mat.int) <- "integer"

# convert matrix back to a df
counts.df.int <- data.frame(counts.mat.int,check.names = F)
```

# Verify that columns of the counts matrix match the rows of the metadata

```{r}

if (all(colnames(counts.df.int) == meta.df$sample.ID)) {
  
  cat("All good.\n")

} else {
  
  cat("The columns of your counts matrix don't match your metadata.\n")
  cat("First 3 samples in counts matrix:", colnames(counts.df.int)[1:3], "\n")
  cat("First 3 samples in metadata:", meta.df$sample.ID[1:3], "\n")

}

```



# Create DESeq2 object
```{r}
dds <- DESeqDataSetFromMatrix(
  countData=counts.df.int, 
  colData=meta.df,
  design= ~ diet * genotype)
```


# Split into pco dco

```{r}
pco.dds <- dds[, data.frame(colData(dds))$location == "pco"]
dco.dds <- dds[, data.frame(colData(dds))$location == "dco"]
slsl.dds <- dds[, data.frame(colData(dds))$genotype == "slsl"]

```

# Filtration
## Remove samples with very low sequencing depth

```{r}
 SAMPLE.SUM.THRESHOLD <- 5000000

data.frame(sample.sum=colSums(counts(dco.dds))) %>%
  rownames_to_column("sample.ID") %>%
  ggplot(aes(y=sample.ID, x=sample.sum)) +
  geom_col() +
  geom_vline(xintercept=SAMPLE.SUM.THRESHOLD, lty=2) +
  labs(title=sprintf("n=%i samples", ncol(dco.dds)))

data.frame(sample.sum=colSums(counts(pco.dds))) %>%
  rownames_to_column("sample.ID") %>%
  ggplot(aes(y=sample.ID, x=sample.sum)) +
  geom_col() +
  geom_vline(xintercept=SAMPLE.SUM.THRESHOLD, lty=2) +
  labs(title=sprintf("n=%i samples", ncol(pco.dds)))

```

## Remove genes with very low expression

```{r}
GENE.SUM.THRESHOLD <- 200

data.frame(row.sum=rowSums(counts(dco.dds))) %>%
  ggplot(aes(log10(row.sum))) +
  geom_histogram() +
  geom_vline(xintercept=log10(GENE.SUM.THRESHOLD), lty=2) +
  labs(title=sprintf("n=%i genes", nrow(dco.dds)))

data.frame(row.sum=rowSums(counts(pco.dds))) %>%
  ggplot(aes(log10(row.sum))) +
  geom_histogram() +
  geom_vline(xintercept=log10(GENE.SUM.THRESHOLD), lty=2) +
  labs(title=sprintf("n=%i genes", nrow(dco.dds)))

```

## Identify mix-ups 
We also identified some potential mixups from mislabeled metadata, so we'll also exclude those here. Reasons for exclusions listed below.

```{r}
dco.mixups <- c("NT90-M-slsl-DCO-chop-uqam-F2-2-45-s-s",
                "NT59-F-sl-sl-1-DCO-2510-chop-uqam-F2-sl-sl",
                "1F-sl-sl-DCO-chop-uqam-p20-sl-sl",
                 "NT88-F-s-s-DCO-chop-uqam-F2-2526-s-s")
      

pco.mixups <- c("NT91-M-slsl-PCO-chop-uqam-F2-2-s-s",
                "1F-sl-sl-PCO-chop-uqam-p20-sl-sl",
                "NT89-F-s-s-PCO-chop-uqam-F2-s-s",
                "NT58-5F-s-s-PCO-2444-chop-uqam-F2-s-s")

slsl.mixups <- c("NT90-M-slsl-DCO-chop-uqam-F2-2-45-s-s",
                "NT59-F-sl-sl-1-DCO-2510-chop-uqam-F2-sl-sl",
                "1F-sl-sl-DCO-chop-uqam-p20-sl-sl",
                 "NT91-M-slsl-PCO-chop-uqam-F2-2-s-s",
                "1F-sl-sl-PCO-chop-uqam-p20-sl-sl")

#DCO
#NT58 (PCO) and NT59(DCO) had the same output, so they were both excluded
#NT90 had inconsistent naming, so it was excluded
#1F and NT88 DCO/PCO suspected mix up

#PCO
#NT91 had inconsistent naming, so it was excluded
#1F and NT89 DCO/PCO suspected mix up

```

##apply filtration and remove mix-ups
```{r}
 dcsamples.to.keep <- colnames(counts(dco.dds))[colSums(counts(dco.dds)) > SAMPLE.SUM.THRESHOLD]
 dcgenes.to.keep <- rownames(counts(dco.dds))[rowSums(counts(dco.dds)) > GENE.SUM.THRESHOLD]
 dcsamples.to.keep2 <- setdiff(dcsamples.to.keep,dco.mixups)
 dco.dds.filt <- dco.dds[dcgenes.to.keep, dcsamples.to.keep2]

 pcsamples.to.keep <- colnames(counts(pco.dds))[colSums(counts(pco.dds)) > SAMPLE.SUM.THRESHOLD]
 pcgenes.to.keep <- rownames(counts(pco.dds))[rowSums(counts(pco.dds)) > GENE.SUM.THRESHOLD]
 pcsamples.to.keep2 <- setdiff(pcsamples.to.keep,pco.mixups)
 pco.dds.filt <- pco.dds[pcgenes.to.keep, pcsamples.to.keep2]
 
 slslsamples.to.keep <- colnames(counts(slsl.dds))[colSums(counts(slsl.dds)) > SAMPLE.SUM.THRESHOLD]
 slslgenes.to.keep <- rownames(counts(slsl.dds))[rowSums(counts(slsl.dds)) > GENE.SUM.THRESHOLD]
 slslsamples.to.keep2 <- setdiff(slslsamples.to.keep, slsl.mixups)
 slsl.dds.filt <- slsl.dds[slslgenes.to.keep, slslsamples.to.keep2]

  dc.slsl.dds.filt <- slsl.dds.filt[, data.frame(colData(slsl.dds.filt))$location == "dco"]
  pc.slsl.dds.filt <- slsl.dds.filt[, data.frame(colData(slsl.dds.filt))$location == "pco"]

 
dc.vsd <- vst(dco.dds.filt, nsub=1000)
pc.vsd <- vst(pco.dds.filt, nsub=1000)
slsl.vsd <- vst(slsl.dds.filt, nsub=1000)
dc.slsl.vsd <- vst(dc.slsl.dds.filt, nsub=1000)
pc.slsl.vsd <- vst(pc.slsl.dds.filt, nsub=1000)
```

#PCA plots 
##All samples with slsl genotype
```{r}

#PCO vs DCO (slsl only)

slsl.p.data <- plotPCA(slsl.vsd, intgroup=c("location", "diet", "genotype"), returnData=TRUE) 

percent.var <- round(100 * attr(slsl.p.data, "percentVar"))


slsl.p <- ggplot(slsl.p.data, aes(x=PC1, y=PC2, color = diet, shape= location)) +

  geom_point(size=3, stroke =2) +

  scale_color_manual(values=c( "darkred","darkgreen")) +

  scale_shape_manual(values=c(0, 2)) +

  coord_fixed() +

  theme(aspect.ratio=1) +

  theme(text=element_text(size=18)) +

  xlab(paste0("PC1: ", percent.var[1], "% variance")) +

  ylab(paste0("PC2: ", percent.var[2], "% variance")) +
  

  ggtitle("Samples with sl/sl genotype")


# pdf("bulkPCA_slslbylocation_Aug2026.pdf", height=6, width=6)
# slsl.p
# dev.off()

```

##Distal colon slsl PCA
```{r}
dc.slsl.p.data <- plotPCA(dc.slsl.vsd, intgroup=c("diet", "genotype"), returnData=TRUE) 

percent.var <- round(100 * attr(dc.slsl.p.data, "percentVar"))


dc.slsl.p <- ggplot(dc.slsl.p.data, aes(x=PC1, y=PC2, color = diet, shape = genotype)) +

  geom_point(size=2, stroke =2) +

  scale_color_manual(values=c( "darkred","darkgreen")) +

  scale_shape_manual(values= 1) +

  coord_fixed() +
  
  stat_ellipse() +

  theme(aspect.ratio=1) +

  theme(text=element_text(size=18)) +

  xlab(paste0("PC1: ", percent.var[1], "% variance")) +

  ylab(paste0("PC2: ", percent.var[2], "% variance")) +
  

  ggtitle("Distal colon sl/sl PCA")


# pdf("bulkPCA_slsl_dco_Aug2026.pdf", height=6, width=6)
# dc.slsl.p
# dev.off()

```

##Proximal colon slsl PCA
```{r}
pc.slsl.p.data <- plotPCA(pc.slsl.vsd, intgroup=c("diet", "genotype"), returnData=TRUE) 

percent.var <- round(100 * attr(pc.slsl.p.data, "percentVar"))


pc.slsl.p <- ggplot(pc.slsl.p.data, aes(x=PC1, y=PC2, color = diet, shape = genotype)) +

  geom_point(size=3, stroke =2) +

  scale_color_manual(values=c( "darkred","darkgreen")) +

  scale_shape_manual(values=1) +

  coord_fixed() +
  
  stat_ellipse() +

  theme(aspect.ratio=1) +

  theme(text=element_text(size=18)) +

  xlab(paste0("PC1: ", percent.var[1], "% variance")) +

  ylab(paste0("PC2: ", percent.var[2], "% variance")) +
  

  ggtitle("Proximal colon sl/sl PCA")


# pdf("bulkPCA_slsl_pco_Aug2026.pdf", height=6, width=6)
# dc.slsl.p
# dev.off()

```

# Find DEGs (differentially expressed genes)


```{r}
dco.dds.filt <- DESeq(dco.dds.filt)
pco.dds.filt <- DESeq(pco.dds.filt)
resultsNames(dco.dds.filt)
resultsNames(pco.dds.filt)

```


## All possible dataframes for comparison

Note: genotypess.dietProt means the default reference condition is slsl and detrimental diet.

### DEGS: slsl only
```{r}
bydiet_slsl.dco.deseq.df<- results(dco.dds.filt, list( c("diet_Prot._vs_Det.") )) %>%
  as.data.frame()

      sig.bydiet_slsl.dco.deseq.df<-bydiet_slsl.dco.deseq.df %>%
      filter(padj<0.05)%>% 
      arrange(log2FoldChange)

      

bydiet_slsl.pco.deseq.df<- results(pco.dds.filt, list( c("diet_Prot._vs_Det.") )) %>%
  as.data.frame()

      sig.bydiet_slsl.pco.deseq.df<-bydiet_slsl.pco.deseq.df %>%
      filter(padj<0.05)%>% 
      arrange(log2FoldChange)
      
#sig.bydiet_slsl.dco.deseq.df are used for pathway analysis and gene ontology
write.table(sig.bydiet_slsl.dco.deseq.df, "dcslsl_dietDEGs.txt",sep="\t")
write.table(sig.bydiet_slsl.pco.deseq.df, "pcslsl_dietDEGs.txt",sep="\t")

```

#Volcano plot
```{r}
#NaN == NaN
```
##DCO volcano
```{r}
PADJ.THRESHOLD <- 0.05 


bydiet_slsl.dco.deseq.df %>%
  rownames_to_column("gene_name") %>% 
  
  #filter missing padj
  dplyr::filter(!is.na(padj)) %>% 
  

  # only label DEGs
  mutate(label=case_when(
    padj < PADJ.THRESHOLD & abs(log2FoldChange)>1.6~ gene_name,
    TRUE ~ NA_character_
  )) %>%

mutate(color.grp = ifelse(log2FoldChange > 0 & -log10(padj) >1.3, yes = "Det.", no = 'color.grp'))%>%
mutate(color.grp = ifelse(log2FoldChange < 0 & -log10(padj) >1.3, yes = "Prot.", no = color.grp))%>%  

ggplot(aes(x=log2FoldChange, y=-log10(padj), label=label)) +
  geom_hline(yintercept=-log10(PADJ.THRESHOLD), lty=2) +
  theme(
    text=element_text(size=18), 
    legend.text=element_text(size=18), 
    axis.text.x=element_text(size=18),
    axis.text.y=element_text(size=18)
  ) +
  geom_text_repel() +
  scale_color_manual(values=c("darkgreen", "lightgray","darkred")) +
  geom_point(aes(colour = colour.grp), alpha=0.1) +
  labs(x="Log2 Fold-change \n<< Higher in Det. group | Higher in Prot. group >>",
       title = "Effect of diet in distal colon in slsl mice")

# pdf("volcano_diet_dco.pdf")
# p
# dev.off()
```
